---
title: "Colors and shapes"
author: "Dr Massimiliano Canzi"
format: 
  revealjs:
    transition-speed: fast
    transition: fade
    background-transition: fade
    chalkboard: true
    slide-number: true
    show-slide-number: print
    touch: true
    controls: true
    theme: theme-mc.scss
    incremental: true
    footer: "`massimiliano.canzi@uni-konstanz.de`, __PRIF__, 01-02.12.2025"
---

```{r}
#| include: false
library(tidyverse)
library(palmerpenguins)
```

## Colors

* _color_
* _colour_

## Colors

There are two main ways of adding color(s): 

* Adding one specific colour
* Assigning colors to dimensions of your data

There's also two types of color patterns: 

* `col` for dots and lines
* `fill` for areas

## Colors 

* If you just want to add a specific color, add `col = "red"` within a specific geometry directly. 
* If you want to assign colors by levels of a variable, specify this in the `aes`

# `col`

## `col` 

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(col = "blue")
```

## `col` 

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, col = species)) + 
  geom_point()
```

## `col` 

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, col = species)) + 
  geom_point(col = "blue") # ??
```

## `col` 

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, col = species)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

## `col`  

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(col = species))
```

## `col` 

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(aes(col = species)) + 
  geom_smooth(method = "lm", col = "black")
```

## `scale_color_manual`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(col = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

# `fill`

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g)) + 
  geom_boxplot()
```

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g)) + 
  geom_boxplot(col = "deepskyblue3") # ??
```

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g)) + 
  geom_boxplot(fill = "deepskyblue3")
```

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g)) + 
  geom_boxplot(fill = "deepskyblue3", alpha = 0.5)
```

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g, fill = species)) + 
  geom_boxplot(alpha = 0.5)
```

## `fill`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = species, y = body_mass_g, fill = species)) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_manual(values = c("indianred2", "deepskyblue3", "goldenrod3"))
```

# `size` and `type`

## `size` and `type`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(col = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

## `size` and `type`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(size = 1) + 
  geom_smooth(method = "lm", aes(col = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

## `size` and `type`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(size = 2) + 
  geom_smooth(method = "lm", aes(col = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

# ??

## `size` and `type`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, size = species)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(col = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

# !!

## `size` and `type`

```{r}
#| echo: true
#| warning: false
penguins %>% 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(aes(pch = species)) + 
  geom_smooth(method = "lm", aes(col = species, linetype = species)) + 
  scale_color_manual(values = c("indianred2", "steelblue3", "goldenrod3"))
```

# Belt and suspenders 

Both keep your trousers where they should be


